This file takes the BEDTools output from `cmds-for-05.sh` and integrates it
with the DESeq2 results as a new column, `gained_ac`. It also adds an
additional `direction` column for up/down genes, and copies the rownames to
a `gene` column.

```{r}
df <- read.table('12_jan_2018/data/deseq-results.txt')
head(df)
```

Use boolean indexing to get the up/down/unchanged genes. There are NAs in this
dataframe (due to the way DESeq2 reports its results) so we also need to get
genes for which the padj is not NA.

```{r}
valid <- !is.na(df$padj)
up <- valid & df$log2FoldChange > 0 & df$padj < 0.1
dn <- valid & df$log2FoldChange < 0 & df$padj < 0.1
df$direction <- 'un'
df$direction[up] <- 'up'
df$direction[dn] <- 'dn'
```

```{r}
hits <- read.table('12_jan_2018/data/tsses_with_gained_h3k27ac.bed')
head(hits)
names(hits) <- c('chrom', 'start', 'stop', 'symbol', 'score', 'strand', 'transcript', 'gene')
head(hits)
```

```{r}
df_idx <- rownames(df) %in% hits$symbol
df$gained_ac <- FALSE
df$gained_ac[df_idx] <- TRUE
head(df)
sum(df$gained_ac)
```


```{r}
df$gene <- rownames(df)
write.table(df, file='bedtools-and-deseq.tsv', sep='\t', quote=FALSE)
```
